---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


Read the data file and convert to a data.table

```{r}
library("data.table")
dat <- read.csv("activity.csv")
dat <- as.data.table(dat)
```

Sum up steps per day

```{r}
# skip NA's for now

good <- complete.cases(dat)
good_dat <- dat[good,]
day_sum <- good_dat[,sum(steps),by="date"]
setnames(day_sum,2,"Total Steps")
```


## What is mean total number of steps taken per day?


Here's a histogram of the total number of steps per day:
```{r, echo=FALSE}
# plot the histogram

br <- c(0,2500,5000,7500,10000,12500,15000,17500,20000,22500)
hist(day_sum$Total,breaks=br,xlab="Total Steps",
     main="Total Steps Taken Each Day")

# Compute mean and median

day_mean <- mean(day_sum$Total)
day_median <- median(day_sum$Total)

```

The mean is `r format(day_mean,nsmall=0)`, the median is `r day_median`


## What is the average daily activity pattern?


Compute the interval averages across all days

```{r}
interval_avg <- good_dat[, mean(steps), by="interval"]
```

Now plot it

```{r}
plot(interval_avg$interval, interval_avg$V1, type='l',
     xlab="Interval",
     ylab="Interval Average",
     main="Mean Steps Taken Each Interval")
```

Calculate the maximum mean steps per interval

```{r}
# Compute mean and median

interval_max <- max(interval_avg$Mean)

```

The maximum average step per interval is `r format(interval_max,digits=5)`


## Imputing missing values


Now count up all of the rows with missing values

```{r}
num_missing <- nrow(dat) - nrow(dat[good,])
```

`r num_missing` rows have NA's

Let's replace each NA with the average for that interval across all days

```{r}
nn <- nrow(dat)
for (j in 1:nn) {
        if(good[j]==FALSE) dat[j,1] <- mean(dat[dat$interval==dat[j,3],3])
}

# make the sums
day_sum_filled <- dat[,sum(steps),by="date"]
setnames(day_sum_filled,2,"Total Steps")
```

Let's look at the new histogram

```{r, echo=FALSE}
hist(day_sum_filled$Total,breaks=br,xlab="Total Steps",
     main="Total Steps Taken Each Day")

# Compute mean and median

day_mean_filled <- mean(day_sum_filled$Total)
day_median_filled <- median(day_sum_filled$Total)
mean_diff <- day_mean - day_mean_filled
median_diff <- day_median - day_median_filled
```

The mean is `r format(day_mean_filled,nsmall=0)`, the median is `r day_median_filled`

The mean has changed by `r mean_diff`, and the median has changed by `r median_diff`. 

The impact of replacing the NA's is to increase the frequency in the first histogram block, which decreases the mean and the median.

## Are there differences in activity patterns between weekdays and weekends?

Let's calculate averages for all weekdays and weekends for each interval

```{r}
days <- as.POSIXlt(dat$date)
#days2 <- character(nrow(dat))
days2 <- as.character(weekdays(days))
nn <- nrow(dat)
day_col <- character(nrow(dat))
for (j in 1:nn) {
        if (days2[j]=="Saturday" | days2[j]=="Sunday") {
                day_col[j] <- "Weekends"
                }        
                else {
                day_col[j] <- "Weekdays"
                }
        }
dat$day <- day_col

interval_avg_2 <- dat[, mean(steps), by=c("interval","day")]
```

Now plot it

```{r}

library(lattice)
## Convert 'Month' to a factor variable

interval_avg_2 <- transform(interval_avg_2, day = factor(day))
xyplot(V1 ~ interval | day, data = interval_avg_2, type="l",layout = c(1, 2))

```
